WORKING TITLE: The adaptive value of green nest material across birds: A systematic review and meta-analysis

### 2025-07-12 manuscript_GNM.qmd
push to continue on laptop

An intriguing nesting behaviour observed in some bird species is the addition of fresh, often aromatic, plant material to the nest. This green nest material (hereafter GNM) --- which often corresponds to leaves, sprigs, herbs and branches--- is typically added from the start of nest building until egg laying, although some species like blue tit *(Cyanistes cyanus)*, keep adding it even after the nest is finished. Importantly, contrary to the structural nest materials (e.g., twigs, branches), GNM is selected from a small, non-random subset of local flora [@clark1985; @petit2002; @garrido-bautista2023], and often characterized by high levels of aromatic volatile compounds (commonly used by songbirds: @clark1985) or secondary metabolites such as oils and resins (commonly used by raptors: @ontiveros2007). GNM is unlikely to be gathered opportunistically, and collecting these uncommon GNM, and replenishing it for some time weeks, requires extra time, and some cases flying outside one's territory [@mennerat2009]. Thus, the general costs associated with nest building [@mainwaring2013], and the additional effort, time and energy devoted to add these non-structural materials suggest an important adaptive function of GNM.

Several hypotheses have been proposed to explain the adaptive function of GNM [@dubiec2013; @scott-baumann2015]. First, the courtship hypothesis suggests that (mostly) males use GNM as visual or olfactory signals to attract females or advertise nest site occupancy to other males [@eens1993; @gwinner1997; @brouwer2004], particularly in cavity-nesting species where visual cues inside the nest are limited. The courtship hypothesis traces back to the early natural history observations by @kessel1957, who noted that male European Starlings (*Sturnus vulgaris*) carried fresh plant material to their nests while singing and displaying to females. Subsequent studies in both European and Spotless Starlings (*Sturnus unicolor*) supported this view that GNM may serve as a sexually selected signal of male quality by showing that females often prefer nests with fresh GNM, and that unpaired males conspicuously deliver GNM during courtship, stopping right after the onset of egg laying [@brouwer2004; @veiga2006], and increasing this behaviour in the presence of potential mates [@eens1993]. The courtship hypothesis was later extended to explain GNM addition solely by females in Blue Tits, therefore suggesting that GNM might serve as a sexually selected signal of female quality [@tomás2013].

Second, the most widely tested hypothesis is the nest protection hypothesis, which proposes that the volatile compounds in GNM reduce the amount of parasites and pathogens in the nests, enhancing offspring survival. This hypothesis traces back to a comparative study in raptors (Falconiformes) that provided indirect evidence for it by comparing nest reuse, diet composition and GNM presence [@wimberger1984]. Soon after, a preference for GNM containing insect-repellent and antimicrobial volatiles was experimentally confirmed in European Starlings, therefore confirming two of the three key criteria for evaluating the nest protection hypothesis (i.e., selective plant use, bioactive chemical properties and demonstrable adaptive benefits; @clark1988). Observational and experimental studies on raptors [@ontiveros2007; @dykstra2009] and passerines [@lafuma2001; @dawson2004; @mennerat2009b; @quiroga2012] have shown mixed results in some cases, GNM was been shown to improve offspring condition without reducing the amount of parasites in the nests [@gwinner2000], which led to the development of the drug hypothesis. This alternative hypothesis suggests that the volatiles present in GNM directly enhance offspring immune function and/or health independent of parasite or pathogen control [@gwinner2000; @mennerat2009a]. Thus, while both the nest protection and drug hypothesis suggest an adaptive benefit of GNM by enhancing offspring survival, the drug hypothesis emphasizes a pharmacological mechanism, whereas the nest protection hypothesis emphasizes parasite and pathogen repellency. Importantly, these two hypotheses are not mutually exclusive unless the improvement in offspring health can be proved to be unrelated to a reduction in parasite or pathogen load, something difficult to achieve.

Last, several additional adaptive hypotheses or functions of GNM that do not refer to its aromatic (i.e., volatiles and secondary metabolites) properties have been also proposed. These include benefits such as enhancing the crypticity of the nest [@skutch1957], providing insulation or sun protection for eggs and offspring [@mertens1977], or simply fulfilling a structural function [@lyons1987]. However, these alternative functional roles have received little support or attention. In all, despite decades of research, abundant observational and experimental studies, and two crucial narrative reviews [@dubiec2013; @scott-baumann2015], we still do not know the overall adaptive significance of GNM, its generality across bird species, and the relative importance of each of the suggested mechanisms. Given that nest building is widespread across taxa ---including but not limited to birds--- and is increasingly recognized as an expression of the extended phenotype, a quantitative synthesis of this field is both timely and has potentially far-reaching implications.

To address this gap, we conducted a pre-registered systematic review and meta-analysis of experiments testing the effect of GNM on various fitness proxies across birds. Our meta-analysis explicitly evaluates the overall evidence for the adaptive significance of GNM, as well as assesses the relative importance of the main suggested hypotheses: the courtship hypothesis and parental care hypothesis (i.e. combination of both the nest protection and drug hypotheses). In addition, we tested several potential moderators that might explain the observed context-dependency (i.e., high heterogeneity among effect sizes). Specifically, we tested whether the effect of GNM varied depending on the type of parasite (arthropods vs. microorganisms), the timing of GNM addition (before hatching, after hatching, or continuously), the experimental design used (aromatic vs. non-aromatic vs. no added material), the focal species studied, and the type of fitness proxy investigated (e.g., physiology, morphology, reproduction). Our meta-analysis of experimental studies is the first quantitative synthesis testing the causal relationship between GNM and fitness across birds.

@@

## Registration and Reporting

We pre-registered our study protocol on the Open Science Framework [@dimri2024] prior to data exploration and analysis. The pre-registration details our objectives, a priori hypotheses and predictions, methods and planned data analyses. Unless stated otherwise, we adhered to the registered protocol. We followed the PRISMA-EcoEvo guidelines [@odea2021] for reporting this study (@figsupp-checklist and @figsupp-prisma-flow). All resources associated with this study, including data and code, are available at \[<https://github.com/shreyadimri/Green_Nest_Material>\] (insert Zenodo citation later).

@@

We conducted a systematic search on Web of Science Core Collection and Scopus on 07.09.2022 (see update below). The keyword search query was designed as a three-block query including terms related to our treatment of interest (("green\*" OR "herb\*" OR "aromatic\*") AND "nest\*"), population of interest ("bird\*" OR "aves" OR "avian" OR "ornithol\*" OR "passerine\*" OR "passeriform\*" OR "songbird\*" OR list of all bird genera) and to limit the search results to experimental studies only ("experiment\*" OR "manipulat\*"). The complete search strings tailored to each specific platform are provided in @figsupp-string. Although our search string only included keywords in English, we planned to screen any study in English, Spanish, Portuguese and Hindi, but found none. We searched through titles, abstracts and keywords with no restrictions on publication date or study type. We validated our search using an initial library of `{r} nrow(own_library)` relevant articles that we had previously identified (@figsupp-own-library), ensuring all were retrieved. Additionally, we used the R package 'litsearchr' v1.0.0 [@grames2019], which did not yield any new relevant terms, confirm our queries. After removing `{r} nrow(full_reference_before_deduplication)-nrow(unique_reference_list)` duplicates using R package 'revtools' v0.4.1 [@westgate2019], we identified `{r} nrow(unique_reference_list)` unique records.

@@

All titles and abstracts were equally allocated and screened independently by atleast two screeners (SD, TR, JMGS) using the R package 'revtools' following a pre-designed decision tree with all inclusion/exclusion criteria (@figsupp-decision-tree). We included all studies on birds that experimentally manipulated the GNM (both aromatics or herbs). `{r} tools::toTitleCase(as.character(english::english(nrow(selected_abstract_screening))))` articles passed the title-and-abstract screening and were further evaluated during the full-text screening, where each article was independently screened by a single screener (SD, TR, JMGS), with another screener verifying the procedure. We used the same criteria as above (@figsupp-decision-tree) with an additional evaluation of whether the variable of interest in the article was in our list of fitness proxies of interest (@figsupp-fitness-proxies). `{r} paste0(toupper(substring(x <- as.character(english::english(nrow(selected_abstract_screening) - nrow(excluded_fulltext))), 1, 1)), substring(x, 2))` passed the full-text screening and were, thus, eligible for our meta-analysis. All conflicts in screening decisions (\~1% disagreement for title-and-abstract screening, and none for full-text screening) were resolved through discussion.

We updated our systematic literature search by repeating our search on the 12th of August, 2024 and found `{r} nrow(unique_reference_list_repeat)` unique articles after removing `{r} nrow(full_reference_before_deduplication_repeat)-nrow(unique_reference_list_repeat)` duplicates. Only one observer (SD) performed both screenings using the same decision trees as before (@figsupp-decision-tree), and found `{r} as.character(english::english(sum(selected_abstract_screening_repeat$screened_abstracts=="selected", na.rm=T), 1, 1))` additional eligible article for our meta-analysis. Lastly, we identified `{r} as.character(english::english(nrow(selected_abstract_screening_repeat)-(sum(selected_abstract_screening_repeat$screened_abstracts=="selected", na.rm=T)), 1, 1))` additional eligible articles recommended by two colleagues. In all, `{r} nrow(selected_abstract_screening)-nrow(excluded_fulltext)+nrow(selected_abstract_screening_repeat)` published articles were eligible for our meta-analysis, but we also obtained two additional unpublished studies (more details below and in @figsupp-prisma-flowchart).

## Data Extraction

@@

The data of each article ware extracted evenly by one of three authors (SD, TR, JMGS), with another screener verifying the procedure, and four authors (SD, TR, JMGS, AST) collectively discussing and revising unclear cases. We prioritised extracting the primary data (i.e., measures of central tendency, variance and sample size) from text and tables, or, if that was not possible, from figures (e.g., boxplots) using the R package `metaDigitise` v.1.0.1 (@pick2018). If primary data were not available, we extracted inferential statistics such as *t* and *F* values (k = `{r} nrow(inferential_only)`effect sizes, n = `{r} length(unique(inferential_only$paper_ID))`) or descriptive statistics such as contingency tables (k = `{r} nrow(contingency_table)`, n = `{r} length(unique(contingency_table$paper_ID))` studies), that could be transformed into an effect size. Last, if the previous steps failed but the article had openly archieved data, we calculated the means, SD and sample sizes from them (n = `{r} length(raw_data)`).

We extracted the means, variances and sample sizes for each control-treatment pairwise comparison. Depending on the experimental design of each study, controls included non-aromatic materials such as grass or moss, lesser amounts of aromatic material than the treatments, no additional material added, or removal of material added by birds; whereas treatments consisted of aromatic materials added by authors, birds, or both. Those comparisons were coded either as non-aromatic GNM vs. aromatic GNM, or no added GNM vs. aromatic GNM, in our moderator "experimental design". However, there was a third type of comparison used in some studies comparing nests with experimentally added non-aromatic GNM to nests without any added GNM or in which GNM added by the birds was removed by the authors. In this type of comparison, the non-aromatic GNM was considered the treatment, and we coded these effect sizes as non-aromatic GNM vs. no added GNM in our moderator "experimental design". The moderator "experimental design" was used for subsequent meta-regressions. For each study, we also extracted information related to the study subjects (species of birds, location of the study population), the methodology of the study (plant species used as treatment and control, fitness proxy measured) and the source from which we extracted data (e.g., figures, tables, raw data files, or supplementary materials). We categorised the fitness proxies measured into 6 broad categories - physiology, morphology, reproduction, behaviour, parasite and pathogenic load, and phenology. For studies testing whether GNM affects the number of parasites or pathogens found in offspring and/or nests, we categorised them into arthropods or microorganisms. For a complete list of variables extracted see @figsupp-variables . We did not include some traits reported in the original studies in all our analyses because it was unclear from the study and our own judgement how they related to fitness (k = `{r} nrow(excluded_proxies)` effect sizes, n = `{r} length(unique(excluded_proxies$paper_ID))`; studies see @figsupp-excluded-proxies for the list and reasons of exclusion).

To obtain missing data (e.g., non-reported results, missing sample sizes), we contacted four authors. All four authors responded to our email, but only two provided the data as requested; one did not have the data any more, and one did not give a conclusive response. Additionally, to locate grey literature (i.e., unpublished data), we wrote to the corresponding author(s) of all the articles that passed full-text screening (n = `{r} (nrow(selected_abstract_screening) - nrow(excluded_fulltext))` studies) to ask for any potential unpublished data that could be included in our meta-analysis. One author provided data for an unpublished study (see description in @figsupp-unpublished-study) and one of us (MO), provided data for another unpublished study. In all, we extracted data for `{r} included_papers_no - 2` published articles and 2 unpublished studies. Author correspondance was performed using a standardised email template (@figsupp-email).

@@

low_sample<-dataset_analysis%>%filter(is.na(SMDH_sign))

dataset_SMDH <- read_csv("data/04_data_analysis/dataset_SMDH.csv", show_col_types = F)
```

We calculated two effect sizes for each pairwise comparison: a log response ratio (lnRR) and a standardised mean difference with heteroskedasticity correction (SMDH), using the `escalc` function (`measure= ROM` and `SMDH` respectively, `vtype= "LS"`) in the R package `metafor` [@metafor]. We specified the treatment group as the numerator and the control group as the denominator for lnRR and treatment minus control for SMDH, such that positive values indicate the 'fitness' value increased in the treatment group. In cases such as measures of parasite load or prevalence, laying or hatching dates, courtship time, scab score etc., the effect size sign was inverted to ensure comparability across all fitness proxies. In all cases, the effect sizes were coded such that the fitness proxies were positively correlated with the overall fitness. SMDH could only be calculated when there were enough data points in both treatment and control groups to estimate the pooled variance--- specifically, when the degrees of freedom (i.e., n1+n2−2) was greater than 1 (excluded k = `{r} nrow(low_sample)`, n = `{r} length(unique(low_sample$paper_ID))`)

For inferential statistics (e.g., *t* value, *F* value), we estimated SMD values (k = `{r} nrow(inferential_only)`, n = `{r} length(unique(inferential_only$paper_ID))`) using the equations provided by @thehand2019. For dichotomous variables (e.g., number of successful nests, k = `{r} nrow(contingency_table)`, n = `{r} length(unique(contingency_table$paper_ID))`), we calculated the log odds ratio (lnOR) using the `escalc` function (`measure=OR`) from `metafor` and subsequently converted it to SMD using equations from @thehand2019. Since lnRR cannot be calculated from inferential statistics, or dichotomous variables, or for comparisons including zero or negative values (k = `{r} nrow(negative_lnRR)`, n = `{r} length(unique(negative_lnRR$paper_ID))`), the number of effect sizes included in the lnRR (k = `{r} nrow(dataset_lnRR)`, n = `{r} length(unique(dataset_lnRR$paper_ID))`) subset is smaller than that of SMDH (k = `{r} nrow(dataset_SMDH)`, n = `{r} length(unique(dataset_SMDH$paper_ID))`).

When calculating each effect size's associated sampling variance, we prioritised the number of nests in each group over the number of individuals as the sample size, whenever possible to reduce the effect of genetic relatedness non-independence. In cases where the control or treatment group was used repeatedly across multiple comparisons, sample sizes were adjusted by dividing them by the number of times each group was used in a comparison to account for shared group non-independence. For the cases where the authors of an original article reported no statistically significant difference between two groups but did not provide numerical values (k = `{r} nrow(no_ES)`, n = `{r} length(unique(no_ES$paper_ID))`), we assigned zero as the effect size for that comparison and imputed its sampling variance using the missing-case approach described in @nakagawa2022 for lnRR and using $\frac{n_1+n_2}{n_1×n_2}$ for SMD [@lajeunesse2013].

## Data Analysis

We performed all analyses in R version 4.5.1 [@R-version]. All multilevel meta-analytical models and meta-regressions were fit using metafor v.4.8-0 [@metafor]. Sampling variances were modelled as variance-covariance matrices assuming a 0.5 correlation (*ρ*) between sampling variances from the same study [@noble2017] using the vcalc function from metafor. (See Supplementary Appendix @figsupp-correlation for sensitivity analyses using different values of *ρ*).

We considered the following levels of non-independence for our analyses: multiple effect sizes collected from (a) same study (Paper ID), (b) bird species (Species ID) and (c) geographical locations (population ID), (d) sub-populations within the same location (Experiment ID; e.g. birds studied at the same location for multiple years or birds studied at the same location separated by distance), (e) groups of individuals within a population (Group ID; eg. males and females in a population, first clutch and second clutch) were represented in our data several times, and (f) repeated measurement of the same trait within the same individuals (Repeated trait ID; e.g. chick mass on day 7, 14 and 21).

Unless stated otherwise, we fitted the following random effects in all our models to account for non-independence: Paper ID, Group ID, Species ID, Population ID and an observation-level random effect (Observation ID). Our random effect structure deviated from our pre-registration since, initially we had pre-registered the inclusion of Experiment ID and repeated trait ID and not population ID. However, Experiment ID and Group ID were almost indistinguishable, so we decided to leave out the one with lesser resolution i.e. Experiment ID. Additionally, few effect sizes were measured repeatedly, making the Repeated trait ID almost identical to the observation-level random effect (Observation ID), so we excluded it (@figsupp-random-effect). Last, we decided to add Population ID, since we noticed that several studies were performed in the same geographical locations.

### Meta-analytic models: biological predictions

>>>>

An intriguing nesting behaviour observed in some bird species is the addition of fresh, often aromatic, plant material to the nest. This green nest material (hereafter GNM), which often corresponds to leaves, sprigs, herbs and branches, is typically added from the start of nest building until egg laying, although some species like the blue tit (Cyanistes cyanus), keep adding it even after the nest is finished. Importantly, contrary to the structural nest materials (e.g., twigs, mud), GNM is selected from a small, non-random subset of local flora [@clark1985; @petit2002; @garrido-bautista2023], and often characterized by high levels of aromatic volatile compounds (commonly used by songbirds: @mennerat2009b) or secondary metabolites such as oils and resins (commonly used by raptors: @ontiveros2007). GNM is unlikely to be gathered opportunistically, and collecting these uncommon GNM, and replenishing it for some times weeks, requires extra time, and in some cases flying outside one’s territory [@mennerat2009]. Thus, the general costs associated with nest-building [@mainwaring2013], and the additional effort, time and energy devoted to add these non-structural materials suggest an important adaptive function of GNM.

Several hypotheses have been proposed to explain the adaptive function of GNM (reviewed in @dubiec2013; @scott-baumann2015) First, the courtship hypothesis suggests that (mostly) males use GNM as visual or olfactory signals to attract females and/or advertise nest site occupancy to other males (e.g., @eens1993; @gwinner1997; @brouwer2004)), particularly in cavity-nesting species where visual cues inside the nest are limited. The courtship hypothesis traces back to the early natural history observations by @kessel1957, who noted that male European Starlings (Sturnus vulgaris) carried fresh plant material to their nests while singing and displaying to females. Subsequent studies in both European and Spotless Starlings (Sturnus unicolor) supported this view that GNM may serve as a sexually selected signal of male quality by showing that females often prefer nests with fresh GNM, and that unpaired males conspicuously deliver GNM during courtship, stopping right after the onset of egg laying [@brouwer2004; @veiga2006], and increasing this behaviour in the presence of potential mates [@eens1993] The courtship hypothesis was later extended to explain GNM addition solely by females in Blue Tits, therefore suggesting that GNM might serve as a sexually selected signal of female quality [@tomás2013].

Second, the most widely tested hypothesis is the nest protection hypothesis, which proposes that the volatile compounds in GNM reduce the amount of parasites and pathogens in the nests, enhancing offspring survival. This hypothesis traces back to a comparative study in raptors (Falconiformes) that provided indirect evidence for it by comparing nest reuse, diet composition and GNM presence [@wimberger1984]. Soon after, a preference for GNM containing insect-repellent and antimicrobial volatiles was experimentally confirmed in European Starlings, therefore confirming two of the three key criteria for evaluating the nest protection hypothesis (i.e., selective plant use, bioactive chemical properties and demonstrable adaptive benefits; Clark and Mason 1988). Observational and experimental studies on raptors [@ontiveros2007; @dykstra2009] and passerines [@lafuma2001; @dawson2004; @mennerat2009b; @quiroga2012] have shown mixed results. In some cases, GNM was shown to improve offspring condition without reducing the amount of parasites in the nests [@gwinner2000], which led to the development of the drug hypothesis. This alternative hypothesis suggests that the volatiles present in GNM directly enhance offspring immune function and/or health independent of parasite or pathogen control [@gwinner2000; @mennerat2009a]. Thus, while both the nest protection and drug hypothesis suggest an adaptive benefit of GNM by enhancing offspring survival, the drug hypothesis emphasizes a pharmacological mechanism, whereas the nest protection hypothesis emphasizes parasite and pathogen repellency. Importantly, these two hypotheses are not mutually exclusive unless the improvement in offspring health can be proved to be unrelated to a reduction in parasite or pathogen load, something difficult to achieve.

Last, several additional adaptive hypotheses or functions of GNM that do not refer to its aromatic (i.e., volatiles and secondary metabolites) properties have been also proposed. These include benefits such as enhancing the crypticity of the nest [@skutch1957], providing insulation or sun protection for eggs and offspring [@mertens1977], or simply fulfilling a structural function [@lyons1987]. However, these alternative functional roles have received little support or attention. In all, despite decades of research, abundant observational and experimental studies, and two crucial narrative reviews [@dubiec2013; @scott-baumann2015], we still do not know the overall adaptive significance of GNM, its generality across bird species, and the relative importance of each of the suggested mechanisms. Given that nest building is widespread across taxa —including but not limited to birds— and is increasingly recognized as an expression of the extended phenotype, a quantitative synthesis of this field is both timely and has potentially far-reaching implications.

To address this gap, we conducted a pre-registered systematic review and meta-analysis of experiments testing the effect of GNM on various fitness proxies across birds. Our meta-analysis explicitly evaluates the overall evidence for the adaptive significance of GNM, as well as assesses the relative importance of the main suggested hypotheses: the courtship hypothesis and parental care hypothesis (i.e., a combination of both the nest protection and drug hypotheses). In addition, we tested several potential moderators that might explain the observed context-dependency (i.e., high heterogeneity among effect sizes). Specifically, we tested whether the effect of GNM varied depending on the experimental design used (aromatic vs. non-aromatic vs. no added material), the timing of GNM addition (before hatching, after hatching, or continuously), the focal species studied, the type of fitness proxy investigated (e.g., physiology, morphology, reproduction), and the type of parasite (arthropods vs. microorganisms). Our meta-analysis of experimental studies is the first quantitative synthesis testing the causal relationship between GNM and fitness across birds.

@@

## Registration and reporting

We pre-registered our study protocol on the Open Science Framework [@dimri2024] prior to data exploration and analysis. The pre-registration details our objectives, a priori hypotheses and predictions, methods and planned data analyses. Unless stated otherwise, we adhered to the registered protocol. We followed the PRISMA-EcoEvo guidelines [@odea2021] for reporting this study (@tabsupp-checklist; @figsupp-prisma-flow). All resources associated with this study, including data and code, are available at \[<https://github.com/shreyadimri/Green_Nest_Material>\] (insert Zenodo citation later).

@@

We conducted a systematic search on Web of Science Core Collection and Scopus on the 7th of September, 2022 (see update below). The keyword search query was designed as a three-block query including terms related to our treatment of interest ((“green\*” OR “herb\*” OR “aromatic\*”) AND “nest\*”), population of interest (“bird\*” OR “aves” OR “avian” OR “ornithol\*” OR “passerine\*” OR “passeriform\*” OR “songbird\*” OR list of all bird genera) and to limit the search results to experimental studies only (“experiment\*” OR “manipulat\*”). The complete search strings tailored to each specific platform are provided in @figsupp-string. Although our search string only included keywords in English, we planned to screen any study in English, Spanish, Portuguese and Hindi, but found none. We searched through titles, abstracts and keywords with no restrictions on publication date or study type. We validated our search using an initial library of 15 relevant articles that we had previously identified (@figsupp-own-library), ensuring all were retrieved. Additionally, we used the R package ‘litsearchr’ v1.0.0 [@grames2019], which did not yield any new relevant terms, confirm our queries. After removing `{r} nrow(full_reference_before_deduplication)-nrow(unique_reference_list)` duplicates using the R package ‘revtools’ v0.4.1 [@westgate2019], we identified `{r} nrow(unique_reference_list)` unique records.

@@

All titles and abstracts were equally allocated and screened independently by at least two screeners (SD, TR, JMGS) using the R package ‘revtools’ following a pre-designed decision tree with all inclusion/exclusion criteria (@figsupp-decision-tree). We included all studies on birds that experimentally manipulated the GNM (both aromatics or herbs). `{r} tools::toTitleCase(as.character(english::english(nrow(selected_abstract_screening))))` articles passed the title-and-abstract screening and were further evaluated during the full-text screening, where each article was independently screened by a single screener (SD, TR, JMGS), with another screener verifying the procedure. We used the same criteria as above (@figsupp-decision-tree) with an additional evaluation of whether the variable of interest in the article was in our list of fitness proxies of interest (@figsupp-fitness-proxies). `{r} paste0(toupper(substring(x <- as.character(english::english(nrow(selected_abstract_screening) - nrow(excluded_fulltext))), 1, 1)), substring(x, 2))` articles passed the full-text screening, and were, thus, eligible for our meta-analysis. All conflicts in screening decisions (1% for title-and-abstract screening, and none for full-text screening) were resolved through discussion.

We updated our systematic literature search by repeating our search on the 12th of August, 2024, and found `{r} nrow(unique_reference_list_repeat)` unique articles after removing `{r} nrow(full_reference_before_deduplication_repeat)-nrow(unique_reference_list_repeat)` duplicates. Only one observer (SD) performed the both screenings using the same decision trees as before (@figsupp-decision-tree), and found `{r} as.character(english::english(sum(selected_abstract_screening_repeat$screened_abstracts=="selected", na.rm=T), 1, 1))` additional eligible article for our meta-analysis. Lastly, we identified `{r} as.character(english::english(nrow(selected_abstract_screening_repeat)-(sum(selected_abstract_screening_repeat$screened_abstracts=="selected", na.rm=T)), 1, 1))` additional eligible articles recommended by two colleagues. In all, `{r} nrow(selected_abstract_screening)-nrow(excluded_fulltext)+nrow(selected_abstract_screening_repeat)` published articles were eligible for our meta-analysis, but we also obtained two additional unpublished studies (more details below and in @figsupp-prisma-flowchart).

## Data extraction

@@

dataset_SMDH <- read_csv("data/04_data_analysis/dataset_SMDH.csv", show_col_types = F)

lnRR_comp3<-dataset_lnRR%>%filter(comparision_type==3)

SMDH_comp3<-dataset_SMDH%>%filter(comparision_type==3)

multiple_comparision <- dataset_analysis %>%
group_by(paper_ID) %>%
  summarise(n_types = n_distinct(comparision_type),             
    types   = paste(sort(unique(comparision_type)), collapse = ", "),
    .groups = "drop") %>%
  filter(n_types > 1)
```

The data of each article were extracted evenly by one of three authors (SD, TR,  JMGS), with another screener verifying the procedure, and four authors (SD, TR, JMGS, AST) collectively discussing and revising unclear cases. We prioritised extracting primary data (i.e., measures of central tendency, variance and sample size) from text and tables, or, if that was not possible, from figures (e.g., boxplots) using the R package ‘metaDigitise’ v1.0.1 [@pick2018]. If primary data were not available, we extracted inferential statistics such as *t* and *F* values or descriptive statistics such as contingency tables (see below), that could be transformed into an effect size. Last, if the previous steps failed but the article had openly archived data, we calculated the means, standard deviation and sample sizes from them (n = `{r} length(raw_data)`).

We extracted the means, variances and sample sizes for each control-treatment pairwise comparison. Depending on the experimental design of each study, controls included non-aromatic materials such as grass or moss, lesser amounts of aromatic material than the treatments, no additional material added, or removal of material added by birds; whereas treatments consisted of aromatic materials added by authors, birds, or both. Those comparisons were coded either as non-aromatic GNM vs. aromatic GNM, or no added GNM vs. aromatic GNM, in our moderator “experimental design”. However, there was a third type of comparison used in some studies comparing nests with experimentally added non-aromatic GNM to nests without any added GNM or in which GNM added by the birds was removed by the authors. In this type of comparison, the non-aromatic GNM was considered the treatment, and we coded these `{r} nrow(lnRR_comp3)` and `{r} nrow(SMDH_comp3)` effect sizes, respectively, as non-aromatic GNM vs. no added GNM in our moderator “experimental design”. Importantly, our conclusions were robust to these effect sizes (\@Table SX), so we kept them in our dataset, and the moderator “experimental design” was used for subsequent meta-regressions. In studies where three groups were compared (n = `{r} nrow(multiple_comparision)`), we extracted all possible pairwise comparisons.

For each study, we also extracted information related to the study subjects (species of birds, geographical location of the study population), the methodology of the study (plant species used as treatment and control, fitness proxy measured) and the source from which we extracted data (e.g., figures, tables, raw data files, supplementary materials). We categorised the fitness proxies measured into 6 broad categories: physiology, morphology, reproduction, behaviour, parasite and pathogenic load, and phenology. For studies testing whether GNM affects the number of parasites or pathogens found in offspring and/or nests, we categorised them into arthropods or microorganisms. For a complete list of variables extracted, see @figsupp-variables . We did not include some traits reported in the original studies (k = `{r} nrow(excluded_proxies)` effect sizes, n = `{r} length(unique(excluded_proxies$paper_ID))` studies) in our analyses because it was unclear from the study and our own judgement how they related to fitness (see @figsupp-excluded-proxies).

To obtain missing data (e.g., non-reported results, missing sample sizes), we contacted four corresponding authors. All four authors responded to our email, but only two provided the data as requested; one did not have the data any more, and one did not give a conclusive response. Additionally, to locate grey literature (i.e., unpublished data), we wrote to the corresponding author(s) of all the (n = `{r} (nrow(selected_abstract_screening) - nrow(excluded_fulltext))` articles that passed full-text screening to ask for any potential unpublished data that could be included in our meta-analysis using a standardised email template (@figsupp-email). One author provided data for an unpublished study on Blue Tits (see description in @figsupp-unpublished-study) and one of us (MO), provided data for an unpublished study on Common Buzzards (Buteo buteo).

@@

low_sample<-read_csv("data/03_data_cleaning/low_sample.csv", show_col_types = F)

Using the means, standard deviations and sample sizes for each control-treatment pairwise comparison, we calculated two effect sizes: a log response ratio (lnRR) and a standardised mean difference with heteroskedasticity correction (SMDH), using the escalc function (measure = ROM and SMDH respectively, vtype = "LS") in the R package ‘metafor’ v4.8-0 [@metafor]. We specified the treatment group as the numerator and the control group as the denominator for lnRR, and treatment minus control for SMDH, such that positive values indicate a ‘fitness’ increase in the treatment group. In cases such as measures of parasite load or prevalence, laying or hatching dates, courtship time, or scab score, the effect size sign was inverted prior to analysis to ensure that a positive value consistently meant an increase in fitness regardless of fitness proxy (i.e., to ensure comparability across effect sizes). SMDH could only be calculated when there were enough data points in both treatment and control groups to estimate the pooled variance; specifically, when the degrees of freedom (i.e., n1+n2−2) was greater than 1. There were `{r} as.character(english::english(nrow(low_sample), 1, 1))` effect sizes from `{r} as.character(english::english(length(unique(low_sample$paper_ID)), 1, 1))`) study that did not fulfil that requirement, which we then excluded entirely from the dataset.

For the few cases for which we could not extract means and variances but inferential statistics such as t and F values, we estimated SMD values (k = `{r} nrow(inferential_only)`, n = `{r} length(unique(inferential_only$paper_ID))`) using the equations provided by @thehand2019. For dichotomous variables (e.g., number of successful nests, k = `{r} nrow(contingency_table)`, n = `{r} length(unique(contingency_table$paper_ID))`), we calculated the log odds ratio (lnOR) using the 'escalc' function (measure=OR) from 'metafor' and subsequently converted it to SMD using equations from @thehand2019. Since lnRR cannot be calculated from inferential statistics, or dichotomous variables, or for comparisons including zero or negative values (k = `{r} nrow(negative_lnRR)`), the number of effect sizes included in the lnRR (k = `{r} nrow(dataset_lnRR)`, n = `{r} length(unique(dataset_lnRR$paper_ID))`) subset is smaller than that of SMDH (k = `{r} nrow(dataset_SMDH)`, n = `{r} length(unique(dataset_SMDH$paper_ID))`).

When calculating each effect size’s associated sampling variance, we prioritised the number of nests in each group over the number of individuals as the sample size whenever possible to reduce the effect of genetic relatedness non-independence. In cases where the same control or treatment group was used across multiple comparisons, sample sizes were adjusted prior to calculating effect sizes by dividing them by the number of times each group was used in a comparison to account for shared group non-independence. For cases where the authors of an original article reported a statistically non-significant difference between two groups but without providing numerical values (k = `{r} nrow(no_ES)`, n = `{r} length(unique(no_ES$paper_ID))`), we assigned zero as the effect size for that comparison and imputed its sampling variance using the missing-case approach described in @nakagawa2022 for lnRR and using $\frac{n_1+n_2}{n_1×n_2}$ for SMD [@lajeunesse2013].

## Data analysis

We performed all analyses in R version 4.5.1 [@R-version]. All multilevel meta-analytical models and meta-regressions were fitted using ‘metafor’ v.4.8-0 [@metafor]. Sampling variances were modelled as variance-covariance matrices assuming a 0.5 correlation (ρ) between sampling variances from the same study [@noble2017] using the vcalc function from ‘metafor’ (See Supplementary Appendix @figsupp-correlation for sensitivity analyses using different values of *ρ*).

We originally considered the following levels of non-independence for our analyses: multiple effect sizes collected from  (a) same study (Paper ID), (b) bird species (Species ID) and (c) geographical locations (population ID), (d) sub-populations within the same location (Experiment ID; e.g., birds studied at the same location for multiple years or birds studied at the same location separated by distance), (e) groups of individuals within a population (Group ID; e.g., males and females in a population, first clutch and second clutch) were represented in our data several times, and (f) repeated measurement of the same trait within the same individuals (Repeated trait ID; e.g., chick mass on day 7, 14 and 21).

Unless stated otherwise, we fitted the following random effects in all our models to account for non-independence: Paper ID, Group ID, Species ID, Population ID and an observation-level random effect (Observation ID). Our random effect structure deviated from our pre-registration since, initially, we had pre-registered the inclusion of Experiment ID and Repeated trait ID but not Population ID. However, Experiment ID and Group ID were almost indistinguishable, so we decided to leave out the one with lesser resolution, i.e. Experiment ID. Additionally, since few effect sizes were measured repeatedly, making Repeated trait ID almost identical to the observation-level random effect Observation ID, we excluded it (@figsupp-random-effect). Last, we decided to add Population ID since we noticed after extracting the data but before any analyses that several studies were performed in the same geographical locations.

### Meta-analytic mean and hypotheses

To estimate the effect of GNM on fitness across birds, we used two effect sizes: lnRR and SMD(H). Thus, throughout, we fitted two separate sets of model for each effect size. First, we fitted multilevel intercept-only meta-analytical model to test the overall effect, its uncertainty and heterogeneity. We used the `rma.mv` function from the 'metafor' package using the 'REML' method and a *t* distribution for calculating the test statistics and confidence intervals for the moderators. We also conducted a sensitivity analysis including some traits for which the relationship with fitness is unclear or debated in the literature (e.g., sex ratio, testosterone levels), to test the robustness of our results (@figsupp-MLMA-excluded; k = `{r} nrow(SAonly_proxies)` , n = `{r} length(unique(SAonly_proxies$paper_ID))`). For all intercept-only models, we estimated the total amount of absolute heterogeneity ($σ^2$; the total variation not explained by sampling error), its sources or relative heterogeneity ($I^2$), mean-standardised metric (CVH2) and variance-mean-standardised metric of heterogeneity (M2) following the pluralistic approach suggested by @yang2024a .

To test the relative importance of the two main adaptive hypotheses suggested for GNM, that is, the courtship and the parental care hypothesis, which combines the nest protection and drug hypotheses, we ran a multilevel meta-regression including the moderator ‘hypothesis’ (levels: courtship, parental care, both). For this and all meta-regressions, we estimated $R_{marginal}^2$,which corresponds to the percentage of heterogeneity explained by the moderators [@nakagawa2012]. Unless stated otherwise, all meta-regressions correspond to unimoderator meta-regressions.

### Exploratory meta-regressions
